A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-06 latex/main.tex

A Viterbi algorithm can be used
to find the most likely ``copying path'' of a focal sample
through a reference panel of samples under the LS model.
Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
which was originally developed to build genome-wide human genealogies and
is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
Using this Viterbi algorithm,
sc2ts deterministically infers the copying path of a sample
collected on a certain date through the samples
collected on earlier dates in a given ARG.

The SARS-CoV-2 sequences were determined
from samples collected densely in time.
SARS-CoV-2 phylogenies have low phylogenetic signals
due to high sampling frequency
(such that sampling and transmission outpace nucleotide substitution).
Therefore, most samples can be explained by a small number of mutations.
Leveraging this property of the dataset,
we developed a likelihood-thresholding strategy
to avoid needing to compute the likelihood of
all possible copying paths through an ARG for every single sample.
%%% TODO: This needs to be explained.

[TODO explain Fig S1 and S2]

Furthermore, to avoid performing redundant HMM matching on identical samples,
new samples which were exactly matched to a (sample or non-sample) node in the ARG
were not added immediately.
Instead, the copying paths of these samples were recorded,
and the samples were eventually added back to the ARG in a post-processing step.

>>>>

We use the Viterbi algorithm implemented in tsinfer~\cite{Kelleher2019}
to efficiently find copying paths under this model. We updated this 
implementation to include a ``likelihood threshold'', which allows 
all possible paths through the HMM with likelihood less than a given
value to be considered equally probable. This approach leads to major
gains in likelihood compression, saving memory and reducing computation
time. We then use this likelihood thresholding approach to iteratively
process the samples for a given daily batch, first finding all exact 
matches, then all with 1 mismatch, etc. By setting the appropriate
likelihood threshold for the HMM, we are guaranteed that any HMM
solutions found at this threshold are correct, and can then run the HMM again
with a lower threshold for the remaining samples. Close matches
are found very quickly using this approach, and it is only
samples that are more distant from the current ARG that require
significant processing time and memory. This is reflected in 
the computaional resources reported in 
Figures~\ref{fig:inference_resources},\ref{fig:samples_per_day}.

% A Viterbi algorithm can be used
% to find the most likely ``copying path'' of a focal sample
% through a reference panel of samples under the LS model.
% Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
% which was originally developed to build genome-wide human genealogies and
% is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
% Using this Viterbi algorithm,
% sc2ts deterministically infers the copying path of a sample
% collected on a certain date through the samples
% collected on earlier dates in a given ARG.

% The SARS-CoV-2 sequences were determined
% from samples collected densely in time.
% SARS-CoV-2 phylogenies have low phylogenetic signals
% due to high sampling frequency
% (such that sampling and transmission outpace nucleotide substitution).
% Therefore, most samples can be explained by a small number of mutations.
% Leveraging this property of the dataset,
% we developed a likelihood-thresholding strategy
% to avoid needing to compute the likelihood of
% all possible copying paths through an ARG for every single sample.
% %%% TODO: This needs to be explained.

% [TODO explain Fig S1 and S2]

% Furthermore, to avoid performing redundant HMM matching on identical samples,
% new samples which were exactly matched to a (sample or non-sample) node in the ARG
% were not added immediately.
% Instead, the copying paths of these samples were recorded,
% and the samples were eventually added back to the ARG in a post-processing step.
